---
title: "Exploring QoL model"
author: "Chu-Chang Ku"
date: "2025-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(rstan)
library(lme4)


load("../data/sub_qol.rdata")

## Frequentist approach

training <- dat_qol$training 

ds_nz <- training %>% filter(EQ5D < 1)


clu <- kmeans(ds_nz$EQ5D, 2)

if (clu$centers[1] > clu$centers[2]) {
  ds_nz <- ds_nz %>% 
    mutate(
      cluster = clu$cluster
    )
} else {
  ds_nz <- ds_nz %>% 
    mutate(
      cluster = 3 - clu$cluster
    )
}


ds_c1 <- ds_nz %>% filter(cluster == 1)
ds_c2 <- ds_nz %>% filter(cluster == 2)
ds_zero <- training %>% 
  mutate(
    cluster = (EQ5D >= 1) + 0,
    d15 = (ti < 15 / 365.25),
    d30 = (ti < 30 / 365.25),
    d45 = (ti < 45 / 365.25),
    d60 = (ti < 60 / 365.25)
  )

ds_clu1 <- ds_nz %>% 
  mutate(
    cluster = (cluster == 1) + 0,
    d15 = (ti < 15 / 365.25),
    d30 = (ti < 30 / 365.25),
    d45 = (ti < 45 / 365.25),
    d60 = (ti < 60 / 365.25)
  )

```



# Dynamics of EQ-5D scores

Firstly, we identified there were three clusters in the EQ-5D scores before the end of disutility period. Since there are clear boundary between the clusters, we simply used the K-means to classify the data points by temporal well-off, mild discomfort, and severe discomfort. Within each cluster, we identified models separately and the proportions of the clusters are modelled separately as well. In total, there are four sub-models for the dynamics of EQ-5D scores: (1) the dynamics of mild discomfort, (2) the dynamics of severe discomfort, (3) the probability of temporal well-off, and (4) the probability of severe discomfort within those has QALY-loss.  

```{r, echo=FALSE}
ds_nz %>% 
  mutate(Cluster = as.character(cluster)) %>% 
  ggplot() +
  geom_histogram(aes(x = EQ5D, fill = Cluster), bins = 40) +
  expand_limits(y = 1)
```



## Variables

- **ti**: time in year
- **d'x'**: indicating the time before 'x'; 'x' $\in$ {15, 30, 45, 60} days, 
- **age**: age of rash onset


## Submodels {.tabset}

### Mild discomfort 

#### Function of time
```{r}
fit <- lmer(EQ5D ~ ti + (1 | PID) + (1 | SID), data = ds_c1); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~      (1 | PID) + (1 | SID), data = ds_c1); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~ ti             + (1 | SID), data = ds_c1); print(formula(fit)); print(AIC(fit))
fit <- lm(  EQ5D ~ ti,                         data = ds_c1); print(formula(fit)); print(AIC(fit))

fit <- lmer(EQ5D ~ 1 + (ti | PID) + (ti | SID), data = ds_c1); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~ 1 +            + (ti | SID), data = ds_c1); print(formula(fit)); print(AIC(fit))
fit <- lm(  EQ5D ~ 1,                         data = ds_c1); print(formula(fit)); print(AIC(fit))
```


#### Adding Age
```{r}
fit <- lmer(EQ5D ~            1 + (ti | PID) + (ti | SID), data = ds_c1); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~          Age + (ti | PID) + (ti | SID), data = ds_c1); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~ poly(Age, 2) + (ti | PID) + (ti | SID), data = ds_c1); print(formula(fit)); print(AIC(fit))
```



#### Best fit
```{r}
fit_c1 <- lmer(EQ5D ~ 1 + (ti | PID) + (ti | SID), data = ds_c1)

plot(resid(fit_c1))
abline(h = 0)
```

#### Model vs. data
```{r}
ds_c1 %>% 
  mutate(
    pred = predict(fit_c1, .)
  ) %>% 
  ggplot()  +
  geom_histogram(aes(x = pred, fill = "Sim"), alpha = 0.6, bins = 30) +
  geom_histogram(aes(x = EQ5D, fill = "Data"), alpha = 0.6, bins = 30)
```






### Severe discomfort 
#### Function of time
```{r}
fit <- lmer(EQ5D ~ ti + (1 | PID) + (1 | SID), data = ds_c2); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~      (1 | PID) + (1 | SID), data = ds_c2); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~ ti             + (1 | SID), data = ds_c2); print(formula(fit)); print(AIC(fit))
fit <- lm(  EQ5D ~ ti,                         data = ds_c2); print(formula(fit)); print(AIC(fit))

fit <- lmer(EQ5D ~ 1 + (ti | PID) + (ti | SID), data = ds_c2); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~ 1 +            + (ti | SID), data = ds_c2); print(formula(fit)); print(AIC(fit))
fit <- lm(  EQ5D ~ 1,                         data = ds_c2); print(formula(fit)); print(AIC(fit))
```


#### Adding Age
```{r}
fit <- lmer(EQ5D ~            1 + (ti | PID) + (ti | SID), data = ds_c2); print(formula(fit)); print(AIC(fit))
# fail to converge for a linear age term
# fit <- lmer(EQ5D ~          Age + (ti | PID) + (ti | SID), data = ds_c2); print(formula(fit)); print(AIC(fit))
fit <- lmer(EQ5D ~ poly(Age, 2) + (ti | PID) + (ti | SID), data = ds_c2); print(formula(fit)); print(AIC(fit))
```



#### Best fit
```{r}
fit_c2 <- lmer(EQ5D ~ 1 + (ti | PID) + (ti | SID), data = ds_c2)

plot(resid(fit_c2))
abline(h = 0)
```

#### Model vs. data
```{r}
ds_c2 %>% 
  mutate(
    pred = predict(fit_c2, .)
  ) %>% 
  ggplot()  +
  geom_histogram(aes(x = pred, fill = "Sim"), alpha = 0.6, bins = 30) +
  geom_histogram(aes(x = EQ5D, fill = "Data"), alpha = 0.6, bins = 30)
```

 
### Probability of temporal well

#### Temporal pattern
```{r}
# fit <- glmer(cluster ~ ti + (1 | PID) + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
# fit <- glmer(cluster ~ 1 + (ti | PID) + (ti | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))


fit <- glmer(cluster ~ ti + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
# fit <- glmer(cluster ~ 1 + (ti | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glm(cluster ~ ti, data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glm(cluster ~ 1, data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))


fit <- glmer(cluster ~ ti + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + d30 + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + d30 + d45 + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + d30 + d45 + d60 + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
```


#### Adding Age
```{r}
fit <- glmer(cluster ~ d15 + d30 + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + d30 + Age + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + d30 + poly(Age, 2) + (1 | SID), data = ds_zero, family = binomial); print(formula(fit)); print(AIC(fit))
```

#### Best fit
```{r}
fit_pz <- glmer(cluster ~ d15 + d30 + Age + (1 | SID), data = ds_zero, family = binomial)

plot(resid(fit_pz, type = "deviance"))
```

#### Model vs. data
```{r}

tibble(Age = 70, ti = seq(0, 1, 0.02)) %>% 
  mutate(
    d15 = (ti < 15 / 365.25),
    d30 = (ti < 30 / 365.25),
    d45 = (ti < 45 / 365.25),
    d60 = (ti < 60 / 365.25)
  ) %>% 
  mutate(
    prob = predict(fit_pz, re.form = NA, newdata = ., type = "response")
  ) %>% 
  ggplot() +
  geom_line(aes(y = prob, x = ti, colour = as.character(Age))) +
  geom_line(data = ds_zero %>% 
              mutate(
                tig = cut(ti * 365.25, seq(0, 365, 15), right = F)
              ) %>% 
              group_by(tig) %>% 
              summarise(prop = mean(cluster)) %>% 
              mutate(
                ti = seq(0, 300 - 15, 15) / 365.25
              ), 
            aes(x = ti, y = prop, colour = "Data")) +
  expand_limits(y = 0:1)
```


### Probability of mild discomfort within those has QALY loss 
#### Temporal pattern
```{r}
# fit <- glmer(cluster ~ ti + (1 | PID) + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
# fit <- glmer(cluster ~ 1 + (ti | PID) + (ti | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))

fit <- glmer(cluster ~ ti + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ 1 + (ti | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))

fit <- glm(cluster ~ ti, data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glm(cluster ~ 1, data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))

# fit <- glmer(cluster ~ Age + (ti | PID) + (ti | SID), data = ds_clu1, family = binomial); ; print(formula(fit)); print(AIC(fit))

fit <- glmer(cluster ~ d15 + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + d30 + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + d30 + d45 + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ d15 + d30 + d45 + d60  + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))

```


#### Adding Age
```{r}
fit <- glmer(cluster ~ d15 + d30 + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ Age + d15 + d30 + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
fit <- glmer(cluster ~ poly(Age, 2) + d15 + d30 + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))
```

#### Best fit
```{r}
fit_pc1 <- glmer(cluster ~ poly(Age, 2) + d15 + d30 + (1 | SID), data = ds_clu1, family = binomial); print(formula(fit)); print(AIC(fit))

plot(resid(fit_pc1, type = "deviance"))
```

#### Model vs. data

```{r}

tibble(Age = 70, ti = seq(0, 1, 0.02)) %>% 
  mutate(
    d15 = (ti < 15 / 365.25),
    d30 = (ti < 30 / 365.25)
  ) %>% 
  mutate(
    prob = predict(fit_pc1, re.form = NA, newdata = ., type = "response")
  ) %>% 
  ggplot() +
  geom_line(aes(y = prob, x = ti, colour = as.character(Age))) +
  geom_line(data = ds_clu1 %>% 
              mutate(
                tig = cut(ti * 365.25, seq(0, 365, 15), right = F)
              ) %>% 
              group_by(tig) %>% 
              summarise(prop = mean(cluster)) %>% 
              mutate(
                ti = seq(0, 300 - 30, 15) / 365.25
              ), 
            aes(x = ti, y = prop, colour = "Data")) +
  expand_limits(y = 0:1)
```



### Summary

```{r}
ms_fin <- list()
ms_fin$c1 <- lmer(EQ5D ~ 1 + (ti | PID) + (ti | SID), data = ds_c1)
ms_fin$c2 <- lmer(EQ5D ~ 1 + (ti | PID) + (ti | SID), data = ds_c2)
ms_fin$pz <- glmer(cluster ~ d15 + d30 + Age + (1 | SID), data = ds_zero, family = binomial)
ms_fin$pc1 <-  glmer(cluster ~ poly(Age, 2) + d15 + d30 + (1 | SID), data = ds_clu1, family = binomial)

## reduced model
ms_red <- ms_fin
ms_red$pz <- glm(cluster ~ 1, data = ds_zero, family = binomial)
ms_red$pc1 <-  glm(cluster ~ 1, data = ds_clu1, family = binomial)

gen_pred <- \(ms) {
  crossing(Age = c(60, 70, 80, 90), ti = seq(0, 1, 1 / 100)) %>% 
    mutate(
      d15 = (ti < 15 / 365.25),
      d30 = (ti < 30 / 365.25)
    ) %>% 
    mutate(
      p_z = predict(ms$pz, ., re.form = NA, type = "response"),
      p_c1 = predict(ms$pc1, ., re.form = NA, type = "response") * (1 - p_z),
      p_c2 = 1 - p_z - p_c1,
      q_c1 = predict(ms$c1, ., re.form = NA),
      q_c2 = predict(ms$c2, ., re.form = NA),
      avg = p_c1 * q_c1 + p_c2 * q_c2
    )
}

pred_bind <- bind_rows(
  ms_fin %>% gen_pred %>% mutate(Model = "Final"),
  ms_red %>% gen_pred %>% mutate(Model = "Reduced"),
)
```


#### QALY for first 6 mo.
```{r}
pred_bind %>% 
  filter(ti < 0.5) %>% 
  group_by(Age, Model) %>% 
  summarise(QALY = mean(avg)) %>% 
  pivot_wider(names_from = Model, values_from = QALY)
```


#### QALY for first year 
```{r}
pred_bind %>% 
  group_by(Age, Model) %>% 
  summarise(QALY = mean(avg)) %>% 
  pivot_wider(names_from = Model, values_from = QALY)
```



